---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 6 - Due date 03/26/21"
author: "Benjamin Culberson"
output: pdf_document
geometry: margin=2.54cm
header-includes:
  - \usepackage{enumerate}
  - \usepackage{enumitem}
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp21.Rmd"). Submit this pdf using Sakai.

## Set up

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(lubridate)
library(ggplot2)
library(Kendall)
library(outliers)
library(tidyverse)
library(smooth)

#New package for M9 to assist with tables
#install.packages("kableExtra")
library(kableExtra)
```

## Importing and processing the data set

Consider the data from the file "Net_generation_United_States_all_sectors_monthly.csv". The data corresponds to the monthly net generation from January 2001 to December 2020 by source and is provided by the US Energy Information and Administration. **You will work with the natural gas column only**.

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

### Q1

Import the csv file and create a time series object for natural gas. Make you sure you specify the **start=** and **frequency=** arguments. Plot the time series over time, ACF and PACF.

```{r}
net_generation <- read.csv(
  file="../Data/Net_generation_United_States_all_sectors_monthly.csv",
  header=TRUE,
  skip=3)

#Inspect data
head(net_generation)
nvar <- ncol(net_generation) - 1
nobs <- nrow(net_generation)

net_generation_processed <-
  net_generation %>%
  mutate( Month = my(Month) ) %>% 
  rename( All.Fuels = all.fuels..utility.scale..thousand.megawatthours ) %>% 
  rename( Coal = coal.thousand.megawatthours ) %>% 
  rename( Natural.Gas = natural.gas.thousand.megawatthours ) %>% 
  rename( Nuclear = nuclear.thousand.megawatthours ) %>% 
  rename( Conventional.Hydroelectric = conventional.hydroelectric.thousand.megawatthours ) %>% 
  arrange( Month )

summary(net_generation_processed)

ts_net_generation <- ts(
  net_generation_processed[,2:(nvar+1)],
  start=c(year(net_generation_processed$Month[1]),month(net_generation_processed$Month[1])),
  frequency=12) 

#note that we are only transforming columns with electricity price, not the date columns  
head(ts_net_generation,15)
tail(ts_net_generation,15)

plot(ts_net_generation[,"Natural.Gas"])

#ACF and PACF plots
par(mfrow=c(1,2))
ACF_Plot <- Acf(ts_net_generation[,"Natural.Gas"], lag = 40, plot = TRUE,main="")
PACF_Plot <- Pacf(ts_net_generation[,"Natural.Gas"], lag = 40, plot = TRUE,main="")
par(mfrow=c(1,1))
```

### Q2

Using the $decompose()$ or $stl()$ and the $seasadj()$ functions create a series without the seasonal component, i.e., a deseasonalized natural gas series. Plot the deseasonalized series over time and corresponding ACF and PACF. Compare with the plots obtained in Q1.

```{r}
#Using R decompose function
decompose_natural_gas <- decompose(ts_net_generation[,"Natural.Gas"],"additive")
plot(decompose_natural_gas)

deseasonal_natural_gas <- seasadj(decompose_natural_gas)
plot(deseasonal_natural_gas)
#ACF and PACF plots
par(mfrow=c(1,2))
ACF_Plot <- Acf(deseasonal_natural_gas, lag = 40, plot = TRUE,main="")
PACF_Plot <- Pacf(deseasonal_natural_gas, lag = 40, plot = TRUE,main="")
par(mfrow=c(1,1))

```
The plots in Q2 clearly show the deseasoned data from Q1. The overall trend appears the same but most  the large jumps I would associate with seasonality have been removed (there still are some here or there). The ACF also has the sinusoidal variation removed in the Q2 data (compared to the Q1) and the spikes in the PACF that exist in the Q1 data do not exist in the Q2 data.

## Modeling the seasonally adjusted or deseasonalized series

### Q3

Run the ADF test and Mann Kendall test on the deseasonalized data from Q2. Report and explain the results.

```{r}
print("Results for ADF test/n")
print(adf.test(deseasonal_natural_gas,alternative = "stationary"))

print("Results of Mann Kendall on deseasonalized Natural Gas data")
print(summary(MannKendall(deseasonal_natural_gas)))
```
The ADF test has a null hypothesis that states that the data has a unit root and that its trend will not eventually return to a stationary trend. The alternative hypothesis states that there is not a unit root and is not dependent entirely on the previous observation. This ADF test shows a p-value with 0.01 which means we reject the null hypothesis, suggesting that we have a deterministic trend. 

The Mann Kendall test's null hypothesis states that the data is stationary and its alternate hypothesis states that the data follows a trend. It's two sided p-value of =< 2.22e-16 tells us to reject the null hypothesis of stationary data.


### Q4

Using the plots from Q2 and test results from Q3 identify the ARIMA model parameters $p,d$ and $q$. Note that in this case because you removed the seasonal component prior to identifying the model you don't need to worry about seasonal component. Clearly state your criteria and any additional function in R you might use. DO NOT use the $auto.arima()$ function. You will be evaluated on ability to can read the plots and interpret the test results.

Based on the ACF and PACF plots from Q2 and the test results from Q3, this ARIMA model has p = 1, d = 1, and q = 0. This is an autoregressive model with a deterministic trend. The ADF and Mann Kendall tests show us that this model has a deterministic trend and must be differenced (d=1) and the slow decay of the ACF along with the lag = 1 cutoff on the PACF show that this is an AR model with p = 1. This is not an MA model so q = 0. 

### Q5

Use $Arima()$ from package "forecast" to fit an ARIMA model to your series considering the order estimated in Q4. Should you allow for constants in the model, i.e., $include.mean = TRUE$ or $include.drift=TRUE$. **Print the coefficients** in your report. Hint: use the $cat()$ function to print.

```{r}
ARIMA_NG <- Arima(deseasonal_natural_gas,order=c(1,1,0),seasonal=c(0,0,0),include.mean=TRUE,include.drift=TRUE)
print(ARIMA_NG)
```

### Q6

Now plot the residuals of the ARIMA fit from Q5 along with residuals ACF and PACF on the same window. You may use the $checkresiduals()$ function to automatically generate the three plots. Do the residual series look like a white noise series? Why?

```{r}
checkresiduals(ARIMA_NG)
```
After running the checkresiduals function, these 3 plots appear to show a reasonable residual white noise series. The residual series seems random, normally distributed and the ACF does not appear to show a significant self-correlation.


## Modeling the original series (with seasonality)

### Q7

Repeat Q4-Q6 for the original series (the complete series that has the seasonal component). Note that when you model the seasonal series, you need to specify the seasonal part of the ARIMA model as well, i.e., $P$, $D$ and $Q$.

```{r}
complete_NG<-ts_net_generation[,"Natural.Gas"]

plot(complete_NG)
#ACF and PACF plots
par(mfrow=c(1,2))
ACF_Plot_2 <- Acf(complete_NG, lag = 40, plot = TRUE,main="")
PACF_Plot_2 <- Pacf(complete_NG, lag = 40, plot = TRUE,main="")
par(mfrow=c(1,1))

print("Results for ADF test/n")
print(adf.test(complete_NG,alternative = "stationary"))

print("Results of Mann Kendall on Seasonal Natural Gas data")
print(summary(MannKendall(complete_NG)))

```
After plotting the seasonal trend and its ACF and PACF, and after running the ADF and Mann Kendall tests, the associated SARIMA model has p = 2, d = 1, and q = 0. This is an autoregressive seasonal model with a deterministic trend. The ADF and Mann Kendall tests show us that this model has a deterministic trend and must be differenced both seasonally and non-seasonally (d=D=1, because I know that this has a seasonal trend) and the slow decay of the ACF along with the lag = 2 cutoff on the PACF show that this is an AR model with p = 2. This is not an MA model so q = 0. The small lag spikes in the PACF likely indicate  that P = 1 and the lag spikes in the ACF at lag = 12, 24, 36 indicate that Q = 1 as well. 

```{r}
ARIMA_NG_seasonal <- Arima(complete_NG,order=c(2,1,0),seasonal=c(1,1,1),include.mean=TRUE,include.drift=TRUE)
print(ARIMA_NG_seasonal)
checkresiduals(ARIMA_NG_seasonal)
```
After running the ARIMA model as (2,1,0)x(1,1,1), the residual series does once again appear to be a reasonable white noise series. The residuals appear to be random, somewhat normally distributed, and there does not seem to be any serious self-correlation.

### Q8

Compare the residual series for Q7 and Q6. Can you tell which ARIMA model is better representing the Natural Gas Series? Is that a fair comparison? Explain your response.

After running the two separate models, both seem to have relatively normally distributed residuals with no significant self correlation - they are both reasonably look like white noise series. However, if I had to pick one of these models I would pick the SARIMA (Q7) model. The ACF of the ARIMA model, while showing only a small level of self correlation, does seem to have a pattern. There are noticeable and seemingly recurring spikes throughout the ACF as we move from lag = 1 to lag = 40. Again, these spikes have low values and I'm not that concerned with them, but the SARIMA model only has a single spike at lag = 8 and so I am more comfortable picking the SARIMA model.


## Checking your model with the auto.arima()

**Please** do not change your answers for Q4 and Q7 after you ran the $auto.arima()$. It is **ok** if you didn't get all orders correctly. You will not loose points for not having the correct orders. The intention of the assignment is to walk you to the process and help you figure out what you did wrong (if you did anything wrong!).

### Q9

Use the $auto.arima()$ command on the **deseasonalized series** to let R choose the model parameter for you. What's the order of the best ARIMA model? Does it match what you specified in Q4?

```{r}
auto.arima(deseasonal_natural_gas)
```
I chose ARIMA(1,1,0) while the auto.arima went with ARIMA(1,1,1). Evidently the function identified a moving average component to the model that I did not. I'm not sure why it would select this order. The PACF of the deseasonal data showed a clear cutoff at lag = 1 and a clear slow decay in the ACF. Based on this information I chose to just include an AR process in my model. There may be another reason to choose an MA process that I am missing, but for nowI remain confident in my reasoning.

The ADF and Mann Kendall tests showed a deterministic trend that I decided needed to be differenced. I chose d = 1 for my model and the auto.arima agreed.

### Q10

Use the $auto.arima()$ command on the **original series** to let R choose the model parameters for you. Does it match what you specified in Q7?

```{r}
auto.arima(complete_NG)
```
I chose ARIMA(2,1,0)(1,1,1) while the auto.arima went with ARIMA(1,0,0)(0,1,1)

I chose to difference the equation both seasonally and non-seasonally and while the auto.arima function agreed with my seasonal difference, it did not with my non-seasonal difference. The reason I chose to include both differences is because the non-seasonal data had to be differenced in the previous question, so I felt like just seasonally differencing the data would be insufficient, but I suppose I may have been wrong.

I chose a non-seasonal AR order of 2, because the PACF of the data cuts off at lag = 2, not 1. The auto.arima function does not agree and felt that an order of 1 was better. I'm not sure why. 

The auto.arima function agreed that the spikes in the ACF at lag = 12, 24, etc. made a strong case for the seasonal MA process to have an order of 1. However, the auto.arima function did not agree that the small spikes along the PACF at roughly lag =12, 24, 36 indicate a seasonal autoregressive component to the model. This is somewhat understandable given that the spikes were actually at lag = 13, 25, 37 but I was under the impression that this might have had something to do with p = 2 (which was not the correct value for p).
